---
title: "Arrests 2010 Census Tracts analysis"
output: pdf_document
---

```{r}
library(tidyverse)
library(Matrix)

rm(list = ls())
gc()
```
## Preprocessing
```{r}
# read the data
my.df = read.csv("../../data/core_datasets/arrests/arrests_2010_cta.csv", stringsAsFactors = T)
```

Remove date and NTA variables, convert to factor location, month and KY_CD.
```{r}
my.df$ARREST_DATE = NULL
my.df$nta2020 = NULL
my.df$GeoID = NULL # redundant with geoid

my.df$geoid = factor(my.df$geoid)
my.df$MONTH = factor(my.df$MONTH)

# add NA category
my.df$KY_CD = ifelse(is.na(my.df$KY_CD), "MISSING", my.df$KY_CD)
my.df$KY_CD = factor(my.df$KY_CD)
```


```{r}
str(my.df)
```

Check for NA
```{r}
apply(my.df, 2, function(col) (sum(is.na(col))))
```


A possibility is to get rid of all NAs rows, the portion of deleted rows would be relatively small (of course we're introducing some bias here).
```{r}
nrow(na.omit(my.df)) / nrow(my.df)
```
Other possibilities would be to impute values for numerical variables (using median, mean or more sofisticated methods).
For simplicity we just delete missing values rows.
```{r}
my.df = na.omit(my.df)
```


## Description
Ideally 2010 data are our training set and 2011 data are the test set.

### Variables Description

## Explorative analysis
A reasonable response variable would be the count of crimes

## Models

### Model selection and issues
Here are present both spatial and temporal variables.








